Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2022-03-04 paper.tex
Implement final changes for MNRAS resubmission

\newcommand{\rev}[1]{{\color{red} {#1}}}
\newcommand{\newold}[2]{{#1}{}}
\newcommand{\revcite}[1]{{\color{red} \underline{#1}}}

@@

In addition to our main goal, of particular interest here is
\rev{\sout{also}} the effect of
anisotropic plasma viscosity, which in the following is found to strongly
influence the growth of the flute instability. It is well known that viscosity
in magnetised plasmas (such as those which make up the solar corona) is
anisotropic and strongly dependent on the strength and direction of the local
magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
To take this into account,
\revcite{\citet{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}}
developed a phenomenological model of anisotropic  viscosity
%\rev{\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}}
that captures the main physics of viscosity in the solar corona as outlined in the
analysis of~\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity in
regions of very weak or zero field strength. For brevity, we will refer to this
model of viscosity as ``the switching model''.
In \revcite{\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}}, 
we implemented the switching model as a module for the widely-used general MHD
code Lare3d~\citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the development of the
nonlinear MHD kink instability and the Kelvin-Helmholtz instability. More
generally, the interest in anisotropic viscosity stems from the open question
of which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet of solving
the coronal heating problem. Using scaling laws, it has been suggested that
viscous heating (generated through anisotropic viscosity) can dwarf that of
Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However, due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small scale
instabilities and turbulence is relatively
unknown~\revcite{\citep{klimchukSolvingCoronalHeating2006}}. In addition to directly
heating the plasma, viscosity plays a part in the damping of instabilities and
waves~\citep{rudermanSlowSurfaceWave2000}. It is this effect we are most
interested in here, and it shall be reported that the use of anisotropic
viscosity permits the \rev{growth} of the flute instability, which is otherwise
strongly damped by isotropic viscosity.

@@

The non-dimensionalisation of equations \eqref{eq:MHD} is identical to that
used in the earlier
works by \revcite{\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}}
\rev{\sout{and in reference (Arber et al. 2001) that describes the code Lare3d we use
for numerical solution, see further below}}.
A typical magnetic field strength $B_0$, density $\rho_0$ and length scale
$L_0$ are chosen and the other variables non-dimensionalised appropriately.
Velocity and time are non-dimensionalised using the Alfv\'en speed $u_A = B_0
/ \sqrt{\rho_0 \mu_0}$ and Alfv\'en crossing time $t_A = L_0/u_A$,
respectively. Temperature is non-dimensionalised via $T_0 = u_A^2 \bar{m}
/ k_B$, where $k_B$ is the Boltzmann constant and $\bar{m}$ is the average mass
of ions, here taken to be $\bar{m} = 1.2m_p$ (a mass typical for the solar
corona) where $m_p$ is the proton mass. Dimensional quantities can be recovered
by multiplying the non-dimensional variables by their respective reference
value (e.g. $\vec{B}_{\dim} = B_0 \vec{B}$). The reference values used here are
$B_0 = 5 \times 10^{-3}$ T, $L_0 = 1$ Mm and $\rho_0 = 1.67 \times 10^{-12}
\ \text{kg m}^{-3}$, giving reference values for the Alfv\'en speed $u_A
= 3.45\ \text{Mm s}^{-1}$, Alfv\'en time $t_A =
0.29\ \text{s}$ and temperature $T_0 = 1.73 \times 10^{9}\ K$.

@@

\begin{table*}
  \caption{
%\rev{\sout{Quantitative differences in the observed perturbations between results
%\rev{with} different resistivity values $\eta$. In all measurements the
%anisotropic viscosity model is used except for $k_\text{kink}$ where the
%isotropic viscosity model us used. Measurements times are listed in the main
%text.}}
Quantitative differences in the observed perturbations between results
\rev{with} different \rev{viscosity models and} resistivity values
$\eta$. Measurements times are listed in the main text.}  
\centering
\begin{tabular}{ccccc}
&\multicolumn{2}{c}{$\eta=10^{-4}$}& \multicolumn{2}{c}{$\eta=10^{-3}$} \\
&
Anisotropic &
\rev{Isotropic} &
Anisotropic &
\rev{Isotropic} \\
\midrule
Theoretical average ideal linear growth rate of flute $\gamma$
& 0.88 
& \rev{0.88} 
& 1.73 
& \rev{1.73} \\
Observed growth rate of flute $\gamma$
& 0.69 
& \rev{--} 
& 1.06 
& \rev{1.06}\\
Observed growth rate of kink $\lambda$
& 2.55 
& \rev{2.97} 
& 0.15 
& \rev{0.15}\\
\midrule
Theoretical radius $r_s$ of peak ideal flute growth rate
& 0.125 
& \rev{0.125} 
& 0.163 
& \rev{0.163}\\
\midrule
\rev{Predicted axial wave number $k_{\text{flute}}$}
& \rev{23.74}
& \rev{23.52}
& \rev{17.15}
& \rev{17.60}\\
Observed axial wave number $k_{\text{flute}}$
& \rev{$22.93$}
& \rev{$22.30$}
& \rev{$16.05$}
& \rev{$16.05$}\\
Observed axial wave number $k_{\text{kink}}$
& {4.57}
& \rev{4.41}
& {3.44}
& \rev{3.49}\\
\midrule
\rev{Cumulative viscous heat at $t=50$}
& \rev{$1.64 \times 10^{-2}$}
& \rev{1.28}
& \rev{$2.89 \times 10^{-3}$}
& \rev{0.370} \\
\rev{Cumulative Ohmic heat at $t=50$}
& \rev{3.06}
& \rev{2.75} 
& \rev{6.54} 
& \rev{6.27} \\
\end{tabular}
\label{tab:kink_fluting_params}
\end{table*}

>>>>

\newcommand{\rev}[1]{{{#1}}}
\newcommand{\newold}[2]{{#1}{}}
\newcommand{\revcite}[1]{{#1}}

@@

In addition to our main goal, of particular interest here is
the effect of
anisotropic plasma viscosity, which in the following is found to strongly
influence the growth of the flute instability. It is well known that viscosity
in magnetised plasmas (such as those which make up the solar corona) is
anisotropic and strongly dependent on the strength and direction of the local
magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
To take this into account,
\revcite{\citet{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}}
developed a phenomenological model of anisotropic  viscosity
%\rev{\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}}
that captures the main physics of viscosity in the solar corona as outlined in the
analysis of~\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity in
regions of very weak or zero field strength. For brevity, we will refer to this
model of viscosity as ``the switching model''.
In \revcite{\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}}, 
we implemented the switching model as a module for the widely-used general MHD
code Lare3d~\citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the development of the
nonlinear MHD kink instability and the Kelvin-Helmholtz instability. More
generally, the interest in anisotropic viscosity stems from the open question
of which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet of solving
the coronal heating problem. Using scaling laws, it has been suggested that
viscous heating (generated through anisotropic viscosity) can dwarf that of
Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However, due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small scale
instabilities and turbulence is relatively
unknown~\revcite{\citep{klimchukSolvingCoronalHeating2006}}. In addition to directly
heating the plasma, viscosity plays a part in the damping of instabilities and
waves~\citep{rudermanSlowSurfaceWave2000}. It is this effect we are most
interested in here, and it shall be reported that the use of anisotropic
viscosity permits the \rev{growth} of the flute instability, which is otherwise
strongly damped by isotropic viscosity.

@@

The non-dimensionalisation of equations \eqref{eq:MHD} is identical to that
used in the earlier
works by \revcite{\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}}
.
A typical magnetic field strength $B_0$, density $\rho_0$ and length scale
$L_0$ are chosen and the other variables non-dimensionalised appropriately.
Velocity and time are non-dimensionalised using the Alfv\'en speed $u_A = B_0
/ \sqrt{\rho_0 \mu_0}$ and Alfv\'en crossing time $t_A = L_0/u_A$,
respectively. Temperature is non-dimensionalised via $T_0 = u_A^2 \bar{m}
/ k_B$, where $k_B$ is the Boltzmann constant and $\bar{m}$ is the average mass
of ions, here taken to be $\bar{m} = 1.2m_p$ (a mass typical for the solar
corona) where $m_p$ is the proton mass. Dimensional quantities can be recovered
by multiplying the non-dimensional variables by their respective reference
value (e.g. $\vec{B}_{\dim} = B_0 \vec{B}$). The reference values used here are
$B_0 = 5 \times 10^{-3}$ T, $L_0 = 1$ Mm and $\rho_0 = 1.67 \times 10^{-12}
\ \text{kg m}^{-3}$, giving reference values for the Alfv\'en speed $u_A
= 3.45\ \text{Mm s}^{-1}$, Alfv\'en time $t_A =
0.29\ \text{s}$ and temperature $T_0 = 1.73 \times 10^{9}\ K$.

@@

\begin{table*}
  \caption{
Quantitative differences in the observed perturbations between results
\rev{with} different \rev{viscosity models and} resistivity values
$\eta$. Measurement times are listed in the main text.}  
\centering
\begin{tabular}{ccccc}
&\multicolumn{2}{c}{$\eta=10^{-4}$}& \multicolumn{2}{c}{$\eta=10^{-3}$} \\
&
Anisotropic &
\rev{Isotropic} &
Anisotropic &
\rev{Isotropic} \\
\midrule
Theoretical average ideal linear growth rate of flute $\gamma$
& 0.88 
& \rev{0.88} 
& 1.73 
& \rev{1.73} \\
Observed growth rate of flute $\gamma$
& 0.69 
& \rev{--} 
& 1.06 
& \rev{1.06}\\
Observed growth rate of kink $\lambda$
& 2.55 
& \rev{2.97} 
& 0.15 
& \rev{0.15}\\
\midrule
Theoretical radius $r_s$ of peak ideal flute growth rate
& 0.125 
& \rev{0.125} 
& 0.163 
& \rev{0.163}\\
\midrule
\rev{Predicted axial wave number $k_{\text{flute}}$}
& \rev{23.74}
& \rev{23.52}
& \rev{17.15}
& \rev{17.60}\\
Observed axial wave number $k_{\text{flute}}$
& \rev{$22.93$}
& \rev{$22.30$}
& \rev{$16.05$}
& \rev{$16.05$}\\
Observed axial wave number $k_{\text{kink}}$
& {4.57}
& \rev{4.41}
& {3.44}
& \rev{3.49}\\
\midrule
\rev{Cumulative viscous heat at $t=50$}
& \rev{$1.64 \times 10^{-2}$}
& \rev{1.28}
& \rev{$2.89 \times 10^{-3}$}
& \rev{0.370} \\
\rev{Cumulative Ohmic heat at $t=50$}
& \rev{3.06}
& \rev{2.75} 
& \rev{6.54} 
& \rev{6.27} \\
\end{tabular}
\label{tab:kink_fluting_params}
\end{table*}
